---
title: "Reproducible Research: Peer Assessment 1"
output: html_document
author: Shekhar Pandey
keep_md: true
---

### Set Global options
```{r setoptions, echo=FALSE}
library(knitr)
opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE, results= 'hold')
```

### Loading and preprocessing the data
```{r}
activity <- read.csv("activity.csv", stringsAsFactors = FALSE, na.strings ="NA")
activity$date <- as.Date(activity$date, "%Y-%m-%d")
```

### Get the summary of data, also observe number of NAs in summary
```{r summary}
summary(activity)
```
### Exclude NA records and create a subset of complete records
```{r completeRecords}
activity.complete <- na.omit(activity)
summary(activity.complete)
```
### Total number of steps taken per day
```{r totalsteps}
library(dplyr)
activity.total <- activity.complete %>% group_by(date) %>%
                  summarize(sum(steps))
names(activity.total) <- c("Day", "TotalSteps")
head(activity.total)
```

### Mean and Median of TotalSteps per day
```{r}
me<- mean(activity.total$TotalSteps)
print(me)
med <- median(activity.total$TotalSteps)
print(med)
```

### Make a histogram of the total number of steps per day
```{r makehist}
library(ggplot2)
ggplot(activity.total, aes(TotalSteps)) +
        geom_histogram(fill='light green', binwidth = 1000,colour="white")  +
        labs(title="Histogram of total steps per day",
             y="Frequency", x="Total Steps per day") +
        geom_vline(xintercept = me, col="yellow", size=1) +
        geom_vline(xintercept = med, col="red") +
        geom_text(data=NULL, x=me+100, y=6, label="mean & median")

```

### What is the average daily activity pattern?
```{r}
activity.mean <- activity.complete %>% group_by(interval) %>%
                 summarize(AvgSteps=mean(steps))

maxAvgStep <- activity.mean[which.max(activity.mean$AvgSteps),]

msg <- paste("Max Avg Steps:", trunc(maxAvgStep$AvgSteps), 
             "@ Interval:", trunc(maxAvgStep$interval))

activity.mean %>% ggplot(aes(interval, AvgSteps)) + 
        geom_line(color="blue", size=1) +
        labs(title="Graph of average steps per day",
             y="Average number of Steps", x="Time Interval") +
        scale_x_continuous(breaks = seq(0,2400,100), limits = c(0,2400)) +
        geom_vline(xintercept = maxAvgStep$interval, col="red")+
        geom_text(data=NULL, x=maxAvgStep$interval+100, 
                  y=maxAvgStep$AvgSteps,label=msg)
```


## Imputing missing values
```{r impute}
newActivity <- activity
na_index <- which(is.na(newActivity$steps))
for(i in na_index){
        intrvl <- newActivity[i,"interval"]
        newActivity[i,]$steps <- subset(activity.mean, interval== intrvl)$AvgSteps
}
```
### Display new activity dataframe
```{r}
head(newActivity)
```
### Total number of steps taken per day for new activity dataframe
```{r}
newActivity.total <- newActivity %>%  group_by(date) %>%
        summarize(sum(steps))

names(newActivity.total) <- c("Day", "TotalSteps")

me.new <- mean(newActivity.total$TotalSteps)
med.new <- median(newActivity.total$TotalSteps)

ggplot(newActivity.total, aes(x=TotalSteps)) + 
        geom_histogram(fill="lightgreen",binwidth=1000, color='white') + 
        scale_y_continuous(breaks=seq(0,12,2)) +
        labs(title="Histogram of steps per day",
             x="Total Number of Steps per Day",
             y ="Frequency")+
        geom_vline(xintercept = me.new, size=1, col="yellow") +
        geom_vline(xintercept = med.new, col="red")


```

### Compare histograms of original data and imputed data
```{r}
p1 <- hist(activity.total$TotalSteps, breaks = seq(-250,21500,500), plot = FALSE)
p2 <- hist(newActivity.total$TotalSteps, breaks = seq(-250,21500,500), 
           plot = FALSE)
plot( p1, col=rgb(.1,.1,.1,alpha=0.8), xlim=c(-250,25000), 
      ylim = c(0,12), main = "", xlab="Total Number of Steps per Day")

plot( p2, col=rgb(.2,.2,.2,alpha=0.5), xlim=c(-250,25000), ylim = c(0,12), add=T)

box()

```

### Are there differences in activity patterns between weekdays and weekends?
Add dayofweek and daytype whether its weekday or weekend.
dayofweek is having following values

1. Sunday
2. Monday
3. Wednesday
4. Thursday
5. Friday
6. Saturday


```{r}
library(lubridate)
newActivity <- newActivity %>% mutate(dayofweek=wday(date),
                                      day.type= ifelse(dayofweek != 1 & 
                                                               dayofweek != 7,    "Weekday", ifelse(dayofweek ==1 | dayofweek ==7, "Weeknd",NA)))
```

### compare the average number of steps taken per 5-minute interval for weekdays and weekends.


```{r}
newActivity %>% group_by(day.type, interval) %>%
        summarize(AvgSteps = mean(steps)) %>%
        ggplot(aes(x=interval, y=AvgSteps)) + 
        geom_line(aes(color=day.type)) +
        facet_wrap(~ day.type, nrow=2, ncol=1) +
        labs(x="interval", y="Avg total number of Steps") +
        theme_bw()

```